---
title: "HW5"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

## Read and Modify Data

We'll start with the d1 and d2 tables from the helper code.

```{r}
source("Data/student-merge.R")
mat_tbl <- d1
por_tbl <- d2

full_tbl_1 <- inner_join(mat_tbl, por_tbl, by = c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
```
We receive an unexpected many-to-many relationship warning.

```{r}
full_tbl_2 <- inner_join(mat_tbl, por_tbl, by = setdiff(names(mat_tbl), c("G1", "G2", "G3", "paid", "absences")))

columns_of_interest <- c("reason", "guardian", "activities", "nursery")

full_tbl_2 <- full_tbl_2 |> 
  mutate(across(columns_of_interest, as.factor))
```

## Contingency Tables

```{r}
table(full_tbl_2$reason, useNA = "always")
```

We don't have any missing variables here, and these counts fit neatly into the definition. For example, 85 students chose this school for reputation.

```{r}
subset_tbl <- full_tbl_2 |> 
  filter(reason == "reputation")

table(subset_tbl$guardian, subset_tbl$nursery, useNA = "always")
table(full_tbl_2$guardian, full_tbl_2$nursery, full_tbl_2$reason, useNA = "always")[,, "reputation"]
```

Here we see the same table produced both ways. Among those who chose the school for the reputation, 5 have the father as primary guardian and did not attend nursery school.

```{r}
full_tbl_2 |> 
  group_by(guardian, nursery) |> 
  summarize(count = n()) |> 
  pivot_wider(names_from = nursery, values_from = count)
```

This one looks similar following the pivot. We have 16 total students who didn't attend nursery school with the father as the primary guardian.

## Numeric EDA

We'll use the math table, the Portuguese table, then both filtered.

```{r}
por_tbl |>  
  summarise(across(c("absences", "age", "G3"), list("mean" = mean, "sd" = sd), .names = "{.fn}_{.col}"))
por_tbl |> 
  filter(nursery == "yes") |>  
  summarise(across(c("absences", "age", "G3"), list("mean" = mean, "sd" = sd), .names = "{.fn}_{.col}"))
```

We can see the standard deviation of G3 among Portuguese students who attended nursery school is 3.28.

```{r}
mat_tbl |> 
  group_by(nursery) |> 
  summarise(across(c("absences", "age", "G3"), list("mean" = mean, "sd" = sd), .names = "{.fn}_{.col}"))

mat_tbl |> 
  group_by(nursery, reason) |> 
  summarise(across(c("absences", "age", "G3"), list("mean" = mean, "sd" = sd), .names = "{.fn}_{.col}"))
```

The mean absences among math students who attended nursery school is 5.79. When subset further among students who also chose the school for course options, that number drops to 4.33.

```{r}
por_tbl |> 
  select(age, absences, G1, G2, G3) |> 
  cor()
```

We see that G2 and G3 are very correlated with a value of .92.

```{r}
plot_1_1 <- mat_tbl |> ggplot(aes(x = G3))
plot_1_1 + geom_histogram(aes(fill = nursery), binwidth = 1) + ggtitle("Histogram of G3")

plot_2_1 <- mat_tbl |> ggplot(aes(x = age))
plot_2_1 + geom_histogram(aes(fill = nursery), binwidth = 1) + ggtitle("Histogram of Age")
```

These give us counts in each bin of the values that fall there for G3 and age.

```{r}
plot_1_2 <- mat_tbl |> ggplot(aes(x = G3))
plot_1_2 + geom_density(aes(fill = nursery, alpha = .5)) + ggtitle("Density of G3")

plot_2_2 <- mat_tbl |> ggplot(aes(x = age))
plot_2_2 + geom_density(aes(fill = nursery, alpha = .5)) + ggtitle("Density of Age")
```

This smooths out the histogram, which doesn't work as well for discrete values like age.

```{r}
plot_1_3 <- mat_tbl |> ggplot(aes(x = G3))
plot_1_3 + geom_boxplot(aes(fill = nursery)) + ggtitle("Boxplot of G3")

plot_2_3 <- mat_tbl |> ggplot(aes(x = age))
plot_2_3 + geom_boxplot(aes(fill = nursery)) + ggtitle("Boxplot of Age")
```

This shows us that the "heart" of each distribution is roughly similar in spread.

```{r}
plot_4_1 <- por_tbl |> ggplot(aes(x = G1, y = G3, color = activities))
plot_4_1 + geom_point() + ggtitle("Plot of G3 vs G1")

plot_4_2 <- por_tbl |> ggplot(aes(x = absences, y = G3, color = activities))
plot_4_2 + geom_point() + ggtitle("Plot of G3 vs absences")
```

```{r}
plot_5_1 <- por_tbl |> ggplot(aes(x = G1, y = G3))
plot_5_1 + geom_point() + ggtitle("Plots of G3 vs G1 given nursery") + facet_wrap(~ nursery)

plot_5_2 <- por_tbl |> ggplot(aes(x = absences, y = G3))
plot_5_2 + geom_point() + ggtitle("Plots of G3 vs absences given nursery") + facet_wrap(~ nursery)
```

```{r}
plot_5_1 <- por_tbl |> ggplot(aes(x = G1, y = G3))
plot_5_1 + geom_point() + ggtitle("Plots of G3 vs G1 given nursery and guardian") + facet_wrap(nursery ~ guardian)

plot_5_2 <- por_tbl |> ggplot(aes(x = absences, y = G3))
plot_5_2 + geom_point() + ggtitle("Plots of G3 vs absences given nursery and guardian") + facet_wrap(nursery ~ guardian)
```

These plots let us pull out some trends. For example, in just about any scenario, we tend to see some increase of G3 with G1.